Near instantaneous response of mesophyll conductance to light availability allows shade leaves to lie in wait

### 2015-07-27 manuscript/manuscript.Rmd
added sigletters to both data tables
made equation edits in methods

minor edits in methods and results of manuscript

---
title: Super catchy title of the manuscript goes here
author: ''
date: ''
output:
  word_document:
    reference_docx: manuscriptstyle.docx
  pdf_document: default
  html_document:
    number_sections: yes
csl: global-change-biology.csl
bibliography: references.bib
---

>>>>

---
title: Super catchy title of the manuscript goes here
author: ''
date: ''
output:
  word_document:
    reference_docx: manuscriptstyle.docx
  pdf_document: default
  html_document:
    number_sections: yes
csl: new-phytologist.csl
bibliography: references.bib
---

### 2015-07-27 manuscript/manuscript_wtc3.Rmd
added sigletters to both data tables
made equation edits in methods

minor edits in methods and results of manuscript

---
title: "Are shade leaves lying in wait? How instantaneous responses of mesophyll conductance to light availability in shade leaves impacts canopy carbon gain"
author: ""
date: ""
output:
  word_document:
    reference_docx: manuscriptstyle.docx
  pdf_document: default
  html_document:
    number_sections: yes
csl: new-phytologist.csl
bibliography: references.bib
---

@@

# Introduction
The need to understand and predict carbon uptake in forest ecosystems is crucially important for assessing the impacts of a changing climate. Specifically, this involves accurate upscaling of leaf level measurements to predict whole canopy function. Due to the costs and limitations of efficient light harvesting plants cannot expose all leaves to full sun [@niinemets2010review], making simple upscaling problematic. Incident PPFD declines exponentially with increasing leaf area index, creating a steep light gradient from the canopy top to bottom [@monsi2005factor], and photosynthesis (A) increases with irradiance from shade to sun until biochemical limitation occurs [@evans1995carbon]. Consequently, leaves modify structural and functional properties to efficiently intercept variable light and enhance photosynthetic capacity [@vogelman1996leaves; @valladares2004photosynthetic]. To estimate whole canopy carbon gain it is thus necessary to account for the non-linear response of photosynthesis to light by distinguishing between shaded and sunlit leaves [@de1997simple; @linderson2012up].

\
The distribution of resources required for A, including nitrogen (N) and water, are also partially defined by canopy light gradients. As A has a saturating response with light and a dependence on N, leaf N should be distributed in the upper canopy to maximize canopy carbon gain [@field1983allocating; @field1986photosynthesis; @peltoniemi2012co]. This can lead to changes in chlorophyll per unit N, chlorophyll a:b ratios, electron transport capacity per unit chlorophyll, and ratios of electron transport capacity to Rubisco activity across leaf types [@evans2001photosynthetic]. Sun leaves also experience greater water limitations in the upper canopy, despite effective vascular systems developed for high radiation loads and transpiration [@sellin2008effects; @niinemets2012optimization]. Higher rates of A and stomatal conductance (g~s~) can only be sustained if the hydraulic conductance (K) is also large enough to avoid low leaf water potentials [@hubbard2001stomatal]. Optimal photosynthetic N investment in the upper canopy will be wasted if photosynthetic capacity exceeds hydraulic supply [@niinemets2012optimization], thus K should also be higher in the upper canopy to supply sunlit leaves with sufficient water [@burgess2006hydraulic;@sellin2007effects ;@sellin2008effects].

\
These theoretical distributions of N and water regulate leaf physiological traits that constrain A differently for sun and shade leaves. The balance between the costs and energetic benefits of these leaf traits affecting photosynthetic capacity create trade-offs impacting canopy carbon gain [@givnish1988adaptation]. During A, CO~2~ diffuses from the atmosphere through stomata (g~s~), intercellular air space and into the chloroplast for fixation by Rubisco (mesophyll conductance, g~m~) [@tazoe2011using] and trade-offs exists that constrain each of these diffusion pathways. For optimal leaf carbon gain these pathways should act to minimize the resistance to CO~2~ diffusion while also limiting the intrinsic energy, water and construction costs which diminish the carbon return from A. As these processes should not be considered independent of each other and must be integrated with hydraulic constraints [@griffiths2013you] predicting their behaviour within canopies remains difficult.

\
First, g~s~ should be efficiently distributed within a canopy to utilize supplies of light, N and water to maximize A [@peltoniemi2012co]. •	This is because stomata are hypothesized to exhibit an optimal behaviour to maximize carbon gain while simultaneously minimizing water loss through transpiration [@cowan1977stomatal]. Second, variation in photosynthetic capacity between sun and shade leaves has also been hypothesized to be due to differences in g~m~ [@duursma2006vertical]. Mesophyll conductance can impose limitations on A as large as those associated with g~s~ [@warren2008stand; @ubierna2011estimation]and these limitations can reduce the efficiency of N use in A [@niinemets2007photosynthesis]. Interpreting the coupled responses of both g~s~ and g~m~ to A across sun and shade leaves has major implications for predicting canopy carbon gain, but empirical measurements across tree canopies are still lacking.

\
Additionally, assessing shade leaf behaviour is made difficult with accounting of short term light fluctuations within a canopy, via sunflecks. How shade leaves utilize sunflecks for short term carbon gain depends on the combined response time of gs and gm and the underlying photosynthetic biochemistry acclimated to a low light environment [@pearcy1990sunflecks]. For example, at short timescales the utilization of sunflecks is limited by the slow response of stomatal opening, effectively limiting the maximum assimilation rate that can be achieved [@way2012sunflecks]. Mesophyll conductance has been shown to respond to environmental factors across both long and rapid time scales, possibly faster than gs [@flexas2008mesophyll]. However, anatomical parameters which regulate gm with changing irradiance such as chloroplast surface area [@terashima2006irradiance] and mesophyll thickness [@boardman1977comparative] are not likely adaptable during short light fluctuations. The physiological behaviour of shade leaves to maximize carbon gain must be assessed as both a degree of acclimation to local irradiance and as a potential response to transitory light availability.

@@

\
In this study we use *Eucalyptus tereticornis* Sm. trees, planted in climate controlled whole tree chambers with ambient and elevated temperature (+3&deg;C) treatments, to empirically evaluate the distribution of resources and leaf physiological behaviour of sun and shade leaves.
1.	If whole tree canopies are optimized for carbon gain, then leaf N, K and photosynthetic capacity were predicted to be higher in full sun leaves.
2.	Under comparable leaf VPD, g~s~ and A should exhibit theoretical optimal behaviour across both sun and shade leaves, while g~m~ was predicted to scale positively with photosynthetic capacity.
3.	As shade leaves are constrained by their underlying biochemistry and slow physiological responses, increases in A following sunfleck simulations were not expected to reach levels of full sun leaves.
4.	The effects of climate warming was predicted to be greater in sun than shade leaves, seen as increases in stomatal closure and reductions in leaf carbon gain during summer months.

@@

## Whole tree chamber experimental design
Twelve *Eucalyptus tereticornis* seedlings, chosen from a single local Cumberland plain cohort, were planted in March 2013 into 12 whole-tree chambers (WTC) at the Hawkesbury Forest Experiment site near Richmond, NSW, Australia. Each chamber has a maximum height of 10 m and seedlings were grown for 15 months.  A detailed description of the WTC operation and design is available in [@barton2010whole]. Six chambers were set to match outside ambient air temperatures (AT) while the remaining 6 experienced a constant elevated air temperature treatment (ET, 3&deg;C). The CO~2~ concentration inside the chamber was set to match outside air, tracking diurnal changes. Due to the temperature treatment air humidity was not explicitly controlled. Trees were watered with 70 l weekly from March 2013 to November 2013.  From December 2013 to final harvest trees were watered fortnightly with mean monthly (100 yr) rainfall amount.  In February 2013 half of the chambers (3 each of AT and ET) were subjected to a drought treatment by withholding watering. Due to a limited range of data for the drought treatment only well-watered trees are reported here.

\
Before seedlings were planted into each chamber they were maintained under well watered conditions in 35 l pots and kept inside each chamber. This allowed for seedlings to gain sufficient biomass before planting while also allowing them to acclimate to chamber temperature treatments. Seedlings were planted into each chamber after mean seedling height reached 100 cm.  After 2 months, floors were installed 45 cm from the soil surface.  This enabled chamber fluxes of CO~2~ and H~2~O from the whole tree canopy to be monitored. Gas exchange measurements were initiated in October 2013 when trees had both ample height growth and canopy development for realistic canopy light gradients to be measured.  At this point, trees had a mean height of 34&pm;15.1 and 418&pm;323.1 cm and an estimated leaf area of 3.9&pm;0.1 and 6.2&pm;0.2 m^2^ for ambient and elevated temperature treatments, respectively.

##Leaf gas exchange, online carbon isotope discrimination and mesophyll conductance
Leaf gas exchange measurements were performed monthly through to the final harvest.  Measurements were taken on a representative sun and shade leaf for each tree during each measurement campaign.  The youngest fully expanded leaf from the stem apex was chosen for gas exchange measurements and sun leaves were measured in the upper third of the canopy. In order to minimize any confounding effects of leaf age, shade leaves were measured first in the lower canopy then gradually higher during each campaign.  As shade leaves most likely developed slower this assured that older leaves that exist in the lower canopy were avoided. The nearest leaf on each branch was sampled for measurement of predawn leaf water potential. All leaves were selected and flagged 24 hours prior to initiation of measurements.

\
Prior to gas exchange measurements photosynthetic photon flux density (PPFD) was recorded at the individual leaf level and at the canopy position for each selected leaf.  A hand-held photosynthetically available radiation meter was used to record leaf level PPFD to ensure that chosen leaves were positioned in the desired light environment, either full sun or full shade. A ceptometer (AccuPAR LP-80, Decagon Devices, Pullman, USA) was then used to measure 1 m integrated PPFD at the canopy height of each chosen leaf type. Five ceptometer readings were recorded within the canopy at random locations of the height of each selected leaf.  The mean of these readings was assumed to represent the overall leaf light environment of representative full sun and shade leaves for each tree. All measurements of PPFD and gas exchange were performed on full sun days between 10:00-14:30 h.

\
Leaf level gas exchange was measured with a standard 6 cm^2^ leaf chamber using a portable gas exchange system (LI-6400, LI-COR, Lincoln, NE, USA).  This system was coupled with a tunable diode laser (TDL;TGA100,Campbell Scientific, Inc., Logan, UT, USA) for concurrent measurements of online carbon isotope discrimination. The CO~2~ in the leaf chamber was set at ambient atmospheric [CO~2~] (400 ppm) with a flow rate of 200 $\mu$mols s^-1^.  Two identical gas exchanges systems were run simultaneously, one in each of a randomly chosen WTC for each temperature treatment. This paired design allowed for direct leaf comparisons to be made with simultaneous measurements of similar leaf types. Leaf temperatures were controlled at the current ambient or +3&deg;C chamber air temperature. PPFD in the chamber was set to match the individual light environment of each leaf type (explained above). Sun flecks were then simulated for shade leaves by increasing the chamber PPFD to the light environment of the full sun leaf in the same tree. When shade leaves were exposed to high light they were allowed enough time to equilibrate with chamber conditions until stable (ca. 25 min).

\
Once the chamber environment for leaves was stable the sample and reference gas lines were diverted to the TDL via T-junctions inserted into the reference gas tube and match valve outlet. These gases were dried by passing through napion gas dryers in the respective gas lines, and then ^12^CO~2~ and ^13^CO~2~ concentrations were measured for each gas by the TDL. Reference, sample and 2 calibration gases were run on alternating 80 s loops (20 s each), one for each paired leaf, for a total of 12 min. This allowed for 4-5 measurements per leaf and data were averaged over the last 10 s of reference and samples gases for calculations. The two calibration gases were drawn from compressed air tanks (330 and 740 ppm CO~2~) in order to correct for gain drift of the TDLAS on each measurement cycle. Net CO~2~ assimilation rate, g~s~, transpiration, and intercellular [CO~2~] were auto logged every 15 s for each gas exchange system over the same 12 min interval.

@@

$$\delta = \left(\frac{R_s}{R_{stnd}} -1\right)1000$$

@@

$$\Delta_{obs} = \frac{\xi(\delta^13C_o-\delta^13C_e)}{1000+\delta^13C_e-\xi(\delta^13C_o-\delta^13C_e)}$$

@@

$$\xi = \frac{C_{e}}{C_{e}-C_{o}}$$

$\xi$ is the ratio of the CO~2~ entering the well mixed leaf chamber to the CO~2~ draw down by the leaf.

\
Second, leaf ^13^CO~2~ during C~3~ photosynthesis ($\delta$^13^C) is the resultant discrimination from CO~2~ diffusion from the atmosphere to the site of carboxylation, consisting or a series of fractionation steps described in [@evans1986carbon]. In this experiment, a modified form of this equation presented in [@evans2013temperature] with ternary effect corrections by [@farquhar2012ternary] was used such that:

from Evans & Von Caemmerer
$\Delta$~o~ = $\Delta$~i~ - $\Delta$~gm~ - $\Delta$~e~ - $\Delta$~f~
where $\Delta$~o~ is the observed discrimation and  $\Delta$~i~, $\Delta$~gm~ , $\Delta$~e~ and $\Delta$~f~ are the contritbutions of fractionataion if C~i~=C~c~, g~m~, respiration and photorespiration, respectively. The equations for each are as follows:

$$
\begin{aligned}
\Delta = &\frac{1}{1-t}a'+\frac{1}{1-t}((1+t)b-a')\frac{C_i}{C_a}\\
            &\frac{1+t}{1-t}\left(b-a'-\frac{eR_d}{A+R_d}\right)\frac{A}{g_mC_a}-\\
            &\frac{1+t}{1-t}\left(\frac{eR_D}{(A+R_d)C_a}(C_i-\Gamma_*)\right)- \frac{1+t}{1-t}\left(f\frac{\Gamma_*}{C_a}\right)\\
\end{aligned}
$$

$$
\begin{aligned}
\Delta_i = &\frac{1}{1-t}a'+\frac{1}{1-t}((1+t)b-a')\frac{C_i}{C_a}\\
\end{aligned}
$$

$$
\begin{aligned}
\Delta_gm = &\frac{1+t}{1-t}\left(b-a'-\frac{eR_d}{A+R_d}\right)\frac{A}{g_mC_a}-\\

\end{aligned}
$$

$$
\begin{aligned}
\Delta_e = &\frac{1+t}{1-t}\left(\frac{eR_D}{(A+R_d)C_a}(C_i-\Gamma_*)\right)
\end{aligned}
$$

$$
\begin{aligned}
\Delta = &\frac{1}{1-t}a'+\frac{1}{1-t}((1+t)b-a')\frac{C_i}{C_a}\\
            &\frac{1+t}{1-t}\left(b-a'-\frac{eR_d}{A+R_d}\right)\frac{A}{g_mC_a}-\\
            &\frac{1+t}{1-t}\left(\frac{eR_D}{(A+R_d)C_a}(C_i-\Gamma_*)\right)- \frac{1+t}{1-t}\left(f\frac{\Gamma_*}{C_a}\right)\\
\end{aligned}
$$

$$
\begin{aligned}
\Delta_f = & \frac{1+t}{1-t}\left(f\frac{\Gamma_*}{C_a}\right)\\
\end{aligned}
$$

$$t = \frac{(1+a')E}{2g^t_{ac}}$$

E denotes the transpiration rate and $\text{g}^t_{ac}$ is the total conductance to CO~2~ diffusion to both the boundary layer and stomatal conductance (von Caemmerer 1981). C~a~ and C~i~ are the atmospheric and intercellular partial pressures and $\Gamma_*$ is the compensation point in the absence of mitochondrial respiration in the light (R~d~). In this experiment both $\Gamma_*$ and R~d~ were derived using a standard Arrhenius function with parameters for *Eucalyptus globulis* from [@crous2012light]. The different fractionation factors include; diffusion through water (a~i~, 1.8&permil;), Rubisco carboxylation (b, 29&permil;), the photorespiratory fractionation (f, 16.2&permil;) and the combined fractionation through the boundary layer and the stomata (a'). a' is defined by:

$$a' = \frac{a_b(C_a-C_i)+a(C_s-C_i)}{C_a-C_i}$$

where C~s~ is the CO~2~ partial pressure at the leaf surface, a~b (2.9&permil;) is the fractionation from boundary layer diffusion and a is the fractionation due to diffusion in air~(4.4&permil;) [@evans1986carbon]. The CO~2~ diffusion from the intercellular airspace to the chloroplast, g~m~, is given by its relationship to the CO~2~ assimilation rate (A) by:

$$g_m = \frac{A}{C_i-C_c}$$

where C~c~ is the chloroplast CO~2~ partial pressure. Examples of this approach to measure gas exchange and carbon isotope discrimination are presented in [@evans2013temperature]. Once g~m~ was calculated C~c~ and the drawdown of CO~2~ from the intercellular arispaces to the site of carboxylation were then estimated using eq. 7.

\
Photosynthetic CO~2~ response (AC~i~) curves were developed at 25&deg;C for one sun and shade leaf for each chamber prior to the initiation of the drought treatment. Each AC~i~ curve began at the reference [CO~2~] of 400 $\mu$l l^-1^ and then consisted of 12 additional steps from [CO~2~] of 50 to 1800 $\mu$l l^-1^ at 25&deg;C at saturating light (1800 $\mu$mols m^-1^ s^-1^). From these curves the photosynthetic parameters, J~max~ and Vc~max~, were quantified using the biochemical model of [@farquhar1980biochemical].

@@

\
Predawn ($\Psi_{pd}$) and midday ($\Psi_l$) leaf water potentials were measured for sun and shade leaves during each gas exchange campaign using a PMS 1505D pressure chamber (PMS Instruments, Albany, OR, USA). Predawn leaf water potential on selected leaves for each chamber was measured before sunrise on the same day as gas exchange measurements. Leaves used for gas exchange were immediately sampled for $\Psi_l$ once measurements were completed. All leaves were detached and immediately stored inside foil covered bags before water potential measurements were performed. Leaf water potential and transpiration (E) from gas exchange were then used to calculate leaf specific hydraulic conductance (K) through the equation:

$$K = \frac{E}{\Psi_{pd}~-\Psi_l-hp_wg}$$

where the term Hp~w~g represents the gravitational potential assumed to be minimal in this experiment due to small tree size [@whitehead1998regulation]. Leaf level instantaneous transpiration efficiency (ITE) was calculated as leaf photosynthesis divided by leaf transpiration. The g~1~ parameter was estimated from ITE to VPD response curves by fitting a rearranged optimal stomatal conductance model for ITE [@medlyn2012reconciling] using non-linear resgression, where K=0.5 [see @duursma2013near].

##Data analysis
Differences in experimental parameters to either the warming treatment or leaf type were analysed by one-way analysis of variance in R [@RDevelopmentCoreTeam2011] with chambers as random effects.  Mixed model ANOVAs of A max and leaf chemistry were performed using the nlme package [@Pinheiro2015] in R  and r~2~ values of mixed models were computed as in [@nakagawa2013general]. Confidence intervals (95 %) of mixed effect linear models of leaf photosynthesis as a functions of different physiological parameters were generated using bootstrapping methods with 999 simulations in the lme4 package [@bates2013lme4] in R. For non-linear mixed models models confidence intervals were estimated by fitting a generalized additive model to the data with the mgcv package in R.  Results were considered significant at P&leq;0.05.

>>>>

---
title: "Near instantaneous response of mesophyll conductance to light availability allows shade leaves to lie in wait"
author: ""
date: ""
output:
  word_document:
    reference_docx: manuscriptstyle.docx
  pdf_document: default
  html_document:
    number_sections: yes
csl: new-phytologist.csl
bibliography: references.bib
---

@@

# Introduction
The need to understand and predict carbon (C) uptake in forest ecosystems is crucially important for assessing the impacts of a changing climate. Specifically, this involves accurate upscaling of leaf level measurements to predict whole canopy function. Due to the costs and limitations of efficient light harvesting plants cannot expose all leaves to full sun [@niinemets2010review], making simple upscaling problematic. Incident PPFD declines exponentially with increasing leaf area index, creating a steep light gradient from the canopy top to bottom [@monsi2005factor], and photosynthesis increases with irradiance from shade to sun until biochemical limitation occurs [@evans1995carbon]. Consequently, leaves modify structural and functional properties to efficiently intercept variable light and enhance photosynthetic capacity [@vogelman1996leaves; @valladares2004photosynthetic]. To estimate whole canopy C gain it is thus necessary to account for the non-linear response of photosynthesis to light by distinguishing between shaded and sunlit leaves [@de1997simple; @linderson2012up].

\
The distribution of resources required for A, including nitrogen (N) and water, are also partially defined by canopy light gradients. As A has a saturating response with light and a dependence on N, leaf N should be distributed in the upper canopy to maximize canopy C gain [@field1983allocating; @field1986photosynthesis; @peltoniemi2012co]. This can lead to changes in chlorophyll per unit N, chlorophyll a:b ratios, electron transport capacity per unit chlorophyll, and ratios of electron transport capacity to Rubisco activity across leaf types [@evans2001photosynthetic]. Sun leaves also experience greater water limitations in the upper canopy, despite effective vascular systems developed for high radiation loads and transpiration [@sellin2008effects; @niinemets2012optimization]. Higher rates of A and stomatal conductance (g~s~) can only be sustained if the hydraulic conductance (K) is also large enough to avoid low leaf water potentials [@hubbard2001stomatal]. Optimal photosynthetic N investment in the upper canopy will be wasted if photosynthetic capacity exceeds hydraulic supply [@niinemets2012optimization], thus K should also be higher in the upper canopy to supply sunlit leaves with sufficient water [@burgess2006hydraulic;@sellin2007effects ;@sellin2008effects].

\
These theoretical distributions of N and water regulate leaf physiological traits that constrain A differently for sun and shade leaves. The balance between the costs and energetic benefits of these leaf traits affecting photosynthetic capacity create trade-offs impacting canopy C gain [@givnish1988adaptation]. During A, CO~2~ diffuses from the atmosphere through stomata (g~s~), intercellular air space and into the chloroplast for fixation by Rubisco (mesophyll conductance, g~m~) [@tazoe2011using] and trade-offs exists that constrain each of these diffusion pathways. For optimal leaf C gain these pathways should act to minimize the resistance to CO~2~ diffusion while also limiting the intrinsic energy, water and construction costs which diminish the C return from A. As these processes should not be considered independent of each other and must be integrated with hydraulic constraints [@griffiths2013you] predicting their behaviour within canopies remains difficult.

\
First, g~s~ should be efficiently distributed within a canopy to utilize supplies of light, N and water to maximize A [@peltoniemi2012co]. •	This is because stomata are hypothesized to exhibit an optimal behaviour to maximize C gain while simultaneously minimizing water loss through transpiration [@cowan1977stomatal]. Second, variation in photosynthetic capacity between sun and shade leaves has also been hypothesized to be due to differences in g~m~ [@duursma2006vertical]. Mesophyll conductance can impose limitations on A as large as those associated with g~s~ [@warren2008stand; @ubierna2011estimation]and these limitations can reduce the efficiency of N use in A [@niinemets2007photosynthesis]. Interpreting the coupled responses of both g~s~ and g~m~ to A across sun and shade leaves has major implications for predicting canopy C gain, but empirical measurements across tree canopies are still lacking.

\
Additionally, assessing shade leaf behaviour is made difficult with accounting of short term light fluctuations within a canopy, via sunflecks. How shade leaves utilize sunflecks for short term C gain depends on the combined response time of gs and gm and the underlying photosynthetic biochemistry acclimated to a low light environment [@pearcy1990sunflecks]. For example, at short timescales the utilization of sunflecks is limited by the slow response of stomatal opening, effectively limiting the maximum assimilation rate that can be achieved [@way2012sunflecks]. Mesophyll conductance has been shown to respond to environmental factors across both long and rapid time scales, possibly faster than gs [@flexas2008mesophyll]. However, anatomical parameters which regulate gm with changing irradiance such as chloroplast surface area [@terashima2006irradiance] and mesophyll thickness [@boardman1977comparative] are not likely adaptable during short light fluctuations. The physiological behaviour of shade leaves to maximize C gain must be assessed as both a degree of acclimation to local irradiance and as a potential response to transitory light availability.

@@

\
In this study we use *Eucalyptus tereticornis* Sm. trees, planted in climate controlled whole tree chambers with ambient and elevated temperature (+3&deg;C) treatments, to empirically evaluate the distribution of resources and leaf physiological behaviour of sun and shade leaves.
1.	If whole tree canopies are optimized for C gain, then leaf N, K and photosynthetic capacity were predicted to be higher in full sun leaves.
2.	Under comparable leaf VPD, g~s~ and A should exhibit theoretical optimal behaviour across both sun and shade leaves, while g~m~ was predicted to scale positively with photosynthetic capacity.
3.	As shade leaves are constrained by their underlying biochemistry and slow physiological responses, increases in A following sunfleck simulations were not expected to reach levels of full sun leaves.
4.	The effects of climate warming was predicted to be greater in sun than shade leaves, seen as increases in stomatal closure and reductions in leaf C gain during summer months.

@@

## Whole tree chamber experimental design
Twelve *Eucalyptus tereticornis* seedlings, chosen from a single local Cumberland plain cohort, were planted in March 2013 into 12 whole-tree chambers (WTC) at the Hawkesbury Forest Experiment site near Richmond, NSW, Australia. Each chamber has a maximum height of 10 m and seedlings were grown for 15 months.  A detailed description of the WTC operation and design is available in [@barton2010whole]. Six chambers were set to match outside ambient air temperatures (AT) while the remaining 6 experienced a constant elevated air temperature treatment of +3&deg;C (ET, ). The CO~2~ concentration inside the chamber was set to match outside air, tracking diurnal changes. Due to the temperature treatment air humidity was not explicitly controlled. Trees were watered weekly with 70 L from March 2013 to November 2013.  From December 2013 to final harvest trees were watered fortnightly with the mean monthly (100 yr) rainfall amount.  In February 2013 half of the chambers (3 each of AT and ET) were subjected to a drought treatment by withholding watering. Due to a limited range of data for the drought treatment only well-watered trees are reported here.

\
Before seedlings were planted into each chamber they were maintained under well watered conditions in 35 L pots and kept inside each chamber. This allowed for seedlings to gain sufficient biomass before planting while also allowing them to acclimate to chamber temperature treatments. Seedlings were planted into each chamber after mean seedling height reached 100 cm.  After 2 months, floors were installed 45 cm from the soil surface.  This enabled chamber fluxes of CO~2~ and H~2~O from the whole tree canopy to be monitored. Gas exchange measurements were initiated in October 2013 when trees had both ample height growth and canopy development for realistic canopy light gradients to be measured.  At this point, trees under AT treatment had a mean diameter of 28.2&pm;1.1 mm, height of 348&pm;15.1 cm and an estimated leaf area of 3.9&pm;0.1 m^2^. For ET treatments trees had a mean diameter of 34.1&pm;2.1 mm, height of 418.3&pm;23.1 cm and an estimated leaf area of 6.2&pm;0.2 m^2^.

##Leaf gas exchange, online carbon isotope discrimination and mesophyll conductance
Leaf gas exchange measurements were performed monthly through to the final harvest.  Measurements were taken on a representative sun and shade leaf for each tree during each measurement campaign.  The newest fully expanded leaf from the stem apex was chosen for gas exchange measurements and sun leaves were measured in the upper third of the canopy. In order to minimize any confounding effects of leaf age, shade leaves were measured first in the lower canopy then gradually higher during each campaign.  As shade leaves most likely developed slower this assured that older leaves in the lower canopy were avoided. The nearest leaf on each branch was sampled for measurement of predawn leaf water potential. All leaves were selected and flagged 24 hours prior to initiation of measurements.

\
Prior to gas exchange measurements photosynthetic photon flux density (PPFD) was recorded at the individual leaf level and at the canopy position for each selected leaf.  A hand-held photosynthetically available radiation meter was used to record leaf level PPFD to ensure that chosen leaves were positioned in the desired light environment, either full sun or full shade. A ceptometer (AccuPAR LP-80, Decagon Devices, Pullman, USA) was then used to measure 1 m integrated PPFD at the canopy height of each chosen leaf type. Five ceptometer readings were recorded within the canopy at random locations of the height of each selected leaf.  The mean of these readings was assumed to represent the local light environment of full sun and shade leaves for each tree. All measurements of PPFD and gas exchange were performed on full sun days between 10:00-14:30 h.

\
Leaf level gas exchange was measured with a standard 6 cm^2^ leaf chamber using a portable gas exchange system (LI-6400, LI-COR, Lincoln, NE, USA).  This system was coupled with a tunable diode laser (TDL;TGA100,Campbell Scientific, Inc., Logan, UT, USA) for concurrent measurements of online C isotope discrimination. The CO~2~ in the leaf chamber was set at ambient atmospheric [CO~2~] (400 ppm) with a flow rate of 200 $\mu$mols s^-1^.  Two identical gas exchanges systems were run simultaneously, one in each of a randomly chosen WTC for each temperature treatment. Leaf temperatures were controlled at the current AT or ET chamber air temperature. PPFD in the chamber was set to match the individual light environment of each leaf type (explained above). Sun flecks were then simulated for shade leaves by increasing the chamber PPFD to the light environment of the full sun leaf in the same tree. The maximum sunfleck response was then recorded once shade leaves re-stablized in the leaf cuvette (ca. 25 min).

\
Once the chamber environment for leaves was stable, for each leaf measurement, the sample and reference gas lines were diverted to the TDL via T-junctions inserted into the reference gas tube and match valve outlet. These gases were dried by passing through napion gas dryers in the respective gas lines, and then ^12^CO~2~ and ^13^CO~2~ concentrations were measured for each gas by the TDL. Reference, sample and 2 calibration gases were run on alternating 80 s loops (20 s each), one for each paired WTC leaf, for a total of 12 min. This allowed for 4-5 measurements per leaf and data were averaged over the last 10 s of reference and samples gases for calculations. The two calibration gases were drawn from compressed air tanks (330 and 740 ppm CO~2~) in order to correct for gain drift of the TDLAS on each measurement cycle. Photosynthesis, g~s~, transpiration, and intercellular [CO~2~] were auto logged every 15 s for each gas exchange system over the same 12 min interval.

@@

\
(1)
$$\delta = \left(\frac{R_s}{R_{stnd}} -1\right)1000$$

@@

\
(2)
$$\Delta_{obs} = \frac{\xi(\delta^13C_o-\delta^13C_e)}{1000+\delta^13C_e-\xi(\delta^13C_o-\delta^13C_e)}$$

@@

\
(3)
$$\xi = \frac{C_{e}}{C_{e}-C_{o}}$$

and $\xi$ is the ratio of the CO~2~ entering the well mixed leaf cuvette to the CO~2~ draw down by the leaf.

Second, C isotope discrimination during C~3~ photosynthesis ($\Delta$) is the resultant discrimination from CO~2~ diffusion from the atmosphere to the site of carboxylation, consisting or a series of fractionation steps described in [@evans1986carbon]. In this experiment, a modified form of this equation presented in [@evans2013temperature] with ternary effect corrections by [@farquhar2012ternary] was used such that:

\
(4)
$$\Delta_o = \Delta_i - \Delta_{gm} - \Delta_e - \Delta_f$$

where $\Delta$~o~ is the observed discrimation and  $\Delta$~i~, $\Delta$~gm~ , $\Delta$~e~ and $\Delta$~f~ are the contributions to fractionataion if C~i~ = C~c~, g~m~, respiration and photorespiration, respectively. The equations for each are as follows:

\
(5)
$$
\begin{aligned}
\Delta_i = &\frac{1}{1-t}a'+\frac{1}{1-t}((1+t)b-a')\frac{C_i}{C_a}
\end{aligned}
$$

\
(6)
$$
\begin{aligned}
\Delta_{gm} = &\frac{1+t}{1-t}\left(b-a'-\frac{eR_d}{A+R_d}\right)\frac{A}{g_mC_a}
\end{aligned}
$$

\
(7)
$$
\begin{aligned}
\Delta_e = &\frac{1+t}{1-t}\left(\frac{eR_D}{(A+R_d)C_a}(C_i-\Gamma_*)\right)
\end{aligned}
$$

\
(8)
$$
\begin{aligned}
\Delta_f = & \frac{1+t}{1-t}\left(f\frac{\Gamma_*}{C_a}\right)\\
\end{aligned}
$$

where the different fractionation factors include; diffusion through water (a~i~, 1.8&permil;), Rubisco carboxylation (b, 29&permil;), the photorespiratory fractionation (f, 16.2&permil;) and the combined fractionation through the boundary layer and the stomata (a'). a' is defined by:

\
(9)
$$a' = \frac{a_b(C_a-C_i)+a(C_s-C_i)}{C_a-C_i}$$

where C~s~ is the CO~2~ partial pressure at the leaf surface, a~b (2.9&permil;) is the fractionation from boundary layer diffusion and a is the fractionation due to diffusion in air~(4.4&permil;) [@evans1986carbon]. C~a~ and C~i~ are the atmospheric and intercellular partial pressures and $\Gamma_{*}$  is the compensation point in the absence of mitochondrial respiration in the light (R~d~). In this experiment both $\Gamma_{*}$ and R~d~ were derived using a standard Arrhenius function with parameters for *Eucalyptus globulis* from [@crous2012light]. E denotes the transpiration rate and $\text{g}^t_{ac}$ is the total conductance to CO~2~ diffusion to both the boundary layer and stomatal conductance **(von Caemmerer 1981). The ternary effect corrections (t) are desribed by:

\
(10)
$$t = \frac{(1+a')E}{2g^t_{ac}}$$

The CO~2~ diffusion from the intercellular airspace to the chloroplast, g~m~, is given by its relationship to the CO~2~ assimilation rate (A) by:

\
(11)
$$g_m = \frac{A}{C_i-C_c}$$

where C~c~ is the chloroplast CO~2~ partial pressure. Once g~m~ was calculated C~c~ and the drawdown of CO~2~ from the intercellular arispaces to the site of carboxylation were then estimated using Equation 11. Examples of this approach to measure gas exchange and C isotope discrimination are presented in [@evans2013temperature]. The variation in $\Delta$~o~ between sun and shade leaves and the simulated sunfleck where then compared as a function of C~i~:C~a~.

\
Photosynthetic CO~2~ response (AC~i~) curves were measured at 25 &deg;C for one sun and shade leaf for each chamber prior to the initiation of the drought treatment. Each AC~i~ curve began at the reference [CO~2~] of 400 $\mu$l l^-1^ and then consisted of 12 additional steps from [CO~2~] of 50 to 1800 $\mu$l l^-1^ at 25 &deg;C at saturating light (1800 $\mu$mols m^-1^ s^-1^). From these curves the photosynthetic parameters, J~max~ and Vc~max~, were quantified using the biochemical model of [@farquhar1980biochemical] and fit with the 'plantecophys' package [@Duursma2014] in R [@RDevelopmentCoreTeam2011].

@@

\
Predawn ($\Psi_{pd}$) and midday ($\Psi_l$) leaf water potentials were measured for sun and shade leaves during each gas exchange campaign using a PMS 1505D pressure chamber (PMS Instruments, Albany, OR, USA). Predawn leaf water potential was measured before sunrise on the same day as gas exchange measurements. Leaves used for gas exchange were immediately sampled for $\Psi_l$ once measurements were completed. All leaves were detached and immediately stored inside foil covered bags before water potential measurements were performed. Leaf water potential and transpiration (E) from gas exchange were then used to calculate leaf specific hydraulic conductance (K) through the equation:

\
(12)
$$K = \frac{E}{\Psi_{pd}~-\Psi_l-hp_wg}$$

where the term Hp~w~g represents the gravitational potential assumed to be minimal in this experiment due to small tree size [@whitehead1998regulation]. Leaf level instantaneous transpiration efficiency (ITE) was calculated as leaf photosynthesis divided by transpiration. The g~1~ parameter was estimated from ITE to VPD response curves by fitting a rearranged optimal stomatal conductance model for ITE [@medlyn2012reconciling] using non-linear resgression, where K=0.5 [see @duursma2013near].

##Data analysis
Differences in experimental parameters to either the warming treatment or leaf type were analysed by mixed-effects models in R with WTC as a random effect. Explained variance (R^2^) of mixed models were computed as in  [@nakagawa2013general]. Confidence intervals (95 %) of mixed effect linear models of leaf photosynthesis as a functions of different physiological parameters were generated using bootstrapping methods with 999 simulations in the 'lme4' package [@bates2013lme4] in R. For non-linear mixed models models confidence intervals were estimated by fitting a generalized additive model to the data with the 'mgcv' package in R.  Results were considered significant at P&leq;0.05.
